For Finitary Induction-Induction, \\ Induction is Enough

### 2020-05-03 paper/universaliit.tex
stuff

We construct the theory of IIT signatures in the following steps:
\begin{enumerate}
\item We view the theory of signatures as a type theory, and we define
  its untyped syntax as mutual inductive types together with typing
  judgments given by inductive predicates on the untyped syntax. Then
  the syntax $\I : \SignAlg$ is constructed using those untyped terms
  for which the typing predicate holds.
\item We construct $\ll\blank\rr:\SignMor\,\I\,M$ for arbitrary
  $M : \SignAlg$, by:
  \begin{enumerate}
  \item defining a relation $\blank \sim \blank$ between the well-typed syntax
    and a given model. The idea is that given a syntactic context $\GG$ and a
    semantic context $\model{\GG}$ of the model $M$, we have
    $\GG \sim \model{\GG}$ if and only if $\ll\GG\rr
    = \model{\GG}$, and similarly for types, terms, and substitutions;
    % semantic context $\model{\GG}$ of
    % the model $M$ if and only if $\GG$ relates to $\rec\,{\GG}$;
    % enjoyed by the initial morphism from the
    % syntax to a given model;
  \item showing that this relation is functional.
    \end{enumerate}
  \item Proving the uniqueness of this morphism by showing that any morphism
    $f:\SignMor\,\I\,M$ satisfies the relation. For example, for any syntactic context
    $\GG$ we have $\GG\sim f\,\GG$.
      % TODO donner composantes sur les contextes
\end{enumerate}
The next sections detail each of these steps.

@@

\begin{definition}[Untyped syntax]
The untyped syntax is defined as the following inductive datatype.
\begin{alignat*}{5}
  & \rlap{(1) Substitution calculus} \\
  & \Conu && : \Set \\
  & \Tyu  && : \Set \\
  & \Subu  && : \Set \\
  % & \Subpre  && : \Set \\
  & \Tmu  && : \Set \\
  & \emptyu && : \Conu \\
    & \epsilonu && : \Subu \\
  & \blank\extu\blank && : \Conu\ra\Tyu\ra\Conu \\
  & \blank\consu\blank && : \Subu\ra\Tmu\ra\Subu \\
  & \varu  && : \N \ra \Tmu \\
  & \rlap{(2) Universe} \\
  & \Uu && : \Tyu \\
  & \Elu && : \Tmu\ra\Tyu \\
  & \rlap{(3) Inductive parameters} \\
  & \Piu && : \Tmu\ra\Tyu\ra\Tyu \\
  & \blank\appu\blank && : \Tmu\ra\Tmu\ra \Tmu \\
  & \rlap{(4) Metatheoretic parameters} \\
  & \Pimu && : (T:\Set)\ra(T\ra\Tyu)\ra\Tyu \\
  & \Piiu && : (T:\Set)\ra(T\ra\Tmu)\ra\Tmu \\
  & \blank\appmu \blank && : \Tmu\ra(\alpha:T) \ra\Tmu\\
  & \rlap{(5) Default value} \\
  & \erru && : \Tmu
\end{alignat*}
\end{definition}
Variables are modeled as de Bruijn indices, i.e.\ as natural numbers
pointing to a position in the context. We use the additional default
constructor $\erru:\Tmu$ in case of error (ill-scoped
substitution). The typing judgments will not mention $\erru$.  The
main interest of $\erru$ is that it behaves like a closed term (which
the theory of signatures lacks), in the sense that it is invariant
under substitution. This makes expected equalities about substitution
true even in the ill-typed case, thus reducing the number of
hypotheses for the corresponding lemmas (see
\Cref{lem:exch-laws}).

@@

\begin{definition}[Typing judgements]\label{def:typing_judgments}
  The typing judgments are defined as the following inductive type indexed over the
  untyped syntax:
\begin{alignat*}{5}
  & \rlap{(1) Substitution calculus} \\[0.5em]
  & \blank \vdash && : \Conu\ra \Set \\
  & \blank \vdash \blank  && : \Conu\ra \Tyu\ra\Set \\
  & \blank \vdash \blank \invar \blank  && : \Conu\ra \N \ra \Tyu\ra \Set \\
  & \blank \vdash \blank \in \blank  && : \Conu\ra \Tmu\ra \Tyu\ra \Set \\
  & \blank \vdash \blank \Rightarrow \blank  && : \Conu\ra \Subu \ra \Conu\ra \Set \\
  % & \Subpre  && : \Set \\
  & \emptyw && : \emptyu \vdash \\
  & \epsilonw && : \GG\vdash \epsilonu \Rightarrow \emptyu \\
  & \blank\extw\blank && : ( \GG\vdash )  \ra  (\GG\vdash A)  \ra
  \GG\extu A \vdash \\
  & \consw && :
    (\GD \vdash) \ra
    (\GG\vdash \sigma \Rightarrow\GD)\ra
    ( \GD \vdash A) \ra
    ( \GG \vdash t \in A[\sigma]) \ra
    \GG \vdash t \consu \sigma \Rightarrow \GD\extu A
   \\
  & \varw  && : (\GG \vdash n \invar A) \ra \GG \vdash \varu n \in A \\
  & \vzw  && : (\GG\vdash)\ra(\GG\vdash A)\ra\GG \extu  A \vdash \mathsf{0} \invar \wku \, A \\
  & \vsw  && : (\GG\vdash)\ra(\GG\vdash A)\ra(\GG \vdash n \invar A) \ra (\GG \vdash B) \ra \GG \extu  B
  \vdash \mathsf{S}\,n \invar \wku \, A \\
  & \rlap{(2) Universe} \\[0.5em]
  & \Uw && : (\GG \vdash)\ra \GG \vdash \Uu \\
  & \Elw && : (\GG \vdash)\ra (\GG \vdash a \in \Uu)\ra\GG \vdash \Elu\,a \\
  & \rlap{(3) Inductive parameters} \\[0.5em]
\  & \Piw && :
    (\GG \vdash)\ra (\GG \vdash a \in \Uu)\ra(\GG \extu
  \Elu\,a \vdash B)\ra\GG \vdash \Piu \, a \, B \\
  & \appw && :
    (\GG \vdash)\ra (\GG \vdash a \in \Uu)\ra(\GG \extu
    \Elu\,a \vdash B)
    \\
    & &&
    \ra(\GG \vdash t \in \Piu \, a \, B )
    \ra
    (\GG \vdash u \in \Elu\,a)
    \ra
    \GG \vdash t \appu u \in  B [0 := u] \\
  & \rlap{(4) Inductive parameters} \\[0.5em]
  & \Pimw && :
    (T:\Set)\ra(A : T\ra \Tyu)\ra(\GG \vdash)\ra
    ((t : T)\ra\GG\vdash A\,t) \ra
    \GG \vdash \Pimu \,T\,A
    \\
    & \appmw  && :
    (T:\Set)\ra(A : T\ra \Tyu)\ra(\GG \vdash)\ra
    ((t : T)\ra\GG\vdash A\,t)
    \\
    & &&
    \ra(\GG \vdash t \in \Pimu \,T\,A)
    \ra (u : T) \ra \GG \vdash t\,\appmu\,u \in A\,u
  \end{alignat*}
\end{definition}
There is possibility of redundancy in the arguments of the constructors. Here,
we are ``{paranoid}'' (following nomenclature from xxx), so that we get more inductive hypotheses when performing
recursion.

@@

\subsubsection{Weakening preserves typing}
\label{lem:weaken-typing}
We prove by mutual induction that typing judgments are stable under weakening,
for contexts, types, terms, and substitutions.
% wkTelw : ∀ {Γp}{Ap}(Aw : Γp ⊢ Ap)Δp (Δw : (Γp ^^ Δp) ⊢) → ((Γp ▶p Ap) ^^ wkTel Δp) ⊢
% liftTw : ∀ {Γp}{Ap}(Aw : Γp ⊢ Ap)Δp{Bp}(Bw : (Γp ^^ Δp) ⊢ Bp) → ((Γp ▶p Ap) ^^ wkTel Δp) ⊢ (liftT ∣ Δp ∣ Bp)
\begin{alignat*}{5}
  & \wf{\wkO} && : (\GG \vdash A)\ra(\GG \merge \GD\vdash) \ra
  \GG \extu A \merge \wkO\, \GD \vdash \\
  % liftV (S (n + p)) (liftV n q) ≡ liftV n (liftV (n + p) q)
  & \wf{\wk} && : (\GG \vdash A)\ra(\GG \merge \GD\vdash B) \ra
  \GG \extu A \merge \wkO\,\GD \vdash \wk_{\length{\GD}}\, B  \\
  & \wf{\wk} && : (\GG \vdash A)\ra(\GG \merge \GD\vdash t\in B) \ra
  \GG \extu A \merge \wkO\,\GD \vdash \wk_{\length{\GD}}\, t \in \wk_{\length{\GD}}\,B  \\
  & \wf{\wkO} && : (\GG \vdash A)\ra(\GG \vdash \sigma
  \Rightarrow \GD) \ra
  \GG \extu A \vdash \wkO\, \sigma \Rightarrow \GD
  \end{alignat*}
\subsubsection{Substitution preserves typing}
  We show that judgments are stable under substitution.
\begin{alignat*}{5}
  % Tyw[] : ∀ {Γp}{Ap}(Aw : Γp ⊢ Ap) {Δp}(Δw : Δp ⊢){σp}(σw :  Δp ⊢ σp ⇒ Γp) → Δp ⊢ (Ap [ σp ]T)
  & \wf{[]} && : (\GG \vdash)\ra(\GD \vdash A)\ra
  (\GG \vdash \sigma \Rightarrow \GD) \ra
  \GG \vdash A [ \sigma ]
  \\
  & \wf{[]} && : (\GG \vdash)\ra(\GD \vdash t \in A)\ra
  (\GG \vdash \sigma \Rightarrow \GD) \ra
  \GG \vdash t [\sigma] \in A [ \sigma ]
  \\
  & \wf{[]} && : (\GD \vdash x \invar A)\ra
  (\GG \vdash \sigma \Rightarrow \GD) \ra
  \GG \vdash x [\sigma] \in A [ \sigma ]
  % ∘w : ∀ {Γ} {Δ σ} (σw :  Δ ⊢ σ ⇒ Γ)
  %  {Y}(Yw : Y ⊢) {δ} (δw :  Y ⊢ δ ⇒ Δ) →
  %  Y ⊢  (σ ∘p δ) ⇒ Γ
  \\
  & \wf{{\circ}} && :
  (\GG \vdash) \ra
  (\GG \vdash \sigma \Rightarrow \GD) \ra
  (\GD \vdash \tau \Rightarrow E) \ra
  \GG \vdash \tau \circ \sigma \Rightarrow E
  % keepw : ∀ {Γp}(Γw : Γp ⊢){Δp}(Δw : Δp ⊢){σp}(σw :  Γp ⊢ σp ⇒ Δp) {Ap}(Aw : Δp ⊢ Ap ∈ Up ) → (Γp ▶p (Elp Ap [ σp ]T )) ⊢ (keep σp) ⇒ (Δp ▶p Elp Ap)
  % & \wf{keep-\El} && : (\GG \vdash)\ra(\GD \vdash A)\ra
  % (\GG \vdash \sigma \Rightarrow \GD) \ra
  % \GG \vdash A [ \sigma ]
  % \\
  \end{alignat*}
\subsubsection{Laws for identity substitutions}
We show category and functor laws involving identity substitution for
well-formed types, terms and substitutions.
\begin{alignat*}{5}
  % [idp]V : ∀ {Γ}{A}{x}(xw : Γ ⊢ x ∈v A) → (x [ idp ∣ Γ ∣ ]V) ≡ V x
  & [\idu] && : (\GG \vdash A)\ra A [ \idu\,\GG]  \\
  & [\idu] && : (\GG \vdash x \invar A)\ra x [ \idu\,\GG] = V x \\
  & [\idu] && : (\GG \vdash t \in A)\ra t [ \idu\,\GG] = t \\
  & \idru && : (\GG \vdash \sigma \Rightarrow\GD)\ra \sigma \circ \idu\,\GG = \sigma \\
  & \idlu && : (\GG \vdash \sigma \Rightarrow\GD)\ra \idu\,\GD\circ \sigma = \sigma
  \end{alignat*}
Finally, we show that the identity substitution itself is well-typed:
  \begin{alignat*}{5}
  & \idw && : (\GG \vdash)\ra \GG \vdash \idu\,\GG \Rightarrow \GG
  \end{alignat*}

\subsubsection{Proof irrelevance and unicity of typing}
\label{ss:uniq-typing-types}
A type is a proposition, or proof-irrelevant, if it has at most one inhabitant.
  \[
    \isaprop\,T := (a : T) \ra (a' : T) \ra a = a'
  \]
We show that each of the typing judgments is unique in the
following sense:
\begin{alignat*}{5}
 & \isp{\Conw} && : \isaprop\, (\GG\vdash) \\
 & \isp{\Tyw} && : \isaprop\, (\GG\vdash A) \\
 & \isp{\wf\Var} && : \isaprop\, (\GG\vdash x \invar A) \\
 & \isp{\Tmw} && : \isaprop\, (\GG\vdash t \in A) \\
 & \isp{\wf\Sub} && : \isaprop\, (\GG\vdash \sigma \Rightarrow \GD)
\end{alignat*}
We also show unicity of typing:
\begin{alignat*}{5}
  &
  \Tmw{=}\Ty & &:
  (\GG \vdash t \in A) \ra
  (\GG \vdash t \in B) \ra A = B
  \\
  &
  \wf\Var{=}\Ty & &:
  (\GG \vdash x \invar A) \ra
  (\GG \vdash x \invar B) \ra A = B
  \end{alignat*}
Let us consider for instance the application constructor $\appw$: for a codomain
type $B$ it yields an overall type $C=B[\mathsf{0} := u]$ for an application. Even
if $C$ is known a priori, there may be another $B$ for which $B[\mathsf{0} := u] = C$,
possibly leading to many proofs that $t\appu u$ has type $C$. Unicity of typing
solves this issue, as $B$ is then uniquely determined by the type $\Piu\,A\,B$
of $t$.

@@

 \subsubsection{Syntactic model}
 The syntactic model is obtained by packing the untyped syntax with the typing
 judgments:
% % \subsection{QIIT model}
% Then, we show that they are stable by weakening and substitution.
% Packing the untyped syntax with well-formed typed judgments yields
% a model $S$ of the QIIT. Let us give the most important components:
\begin{alignat*}{5}
 & \syn\Con && := \sum_\GG \GG\vdash
 \\
 & \syn\Ty \,(\GG,\wf\GG)&& := \sum_A \GG\vdash A
 \\
 & \syn\Tm \,(\GG,\wf\GG)(A,\wf A)&& := \sum_t \GG\vdash t \in A
 \\
 & \syn\Sub \,(\GG,\wf\GG)(\GD,\wf \GD)&& := \sum_\sigma \GG\vdash \sigma \Rightarrow \GD
\end{alignat*}
The other fields are given straightforwardly.  Regarding the equations, it is
enough to prove them only for the untyped syntactic part: as we argued in
Section~\ref{ss:uniq-typing-types}, the proofs of typing judgments are
automatically equal.

Up until the construction of the syntactic model, UIP is not used. Function
extensionality on the other hand is necessary because the untyped metatheoretic
$\Pi$ takes a metatheoretic function as an argument. An example induction step
that uses function extensionality is in the type preservation proof for identity substitutions (this
is an equation satisfied by a model of the QIIT), in particular in the case
$(\Pim\,T\, A)[\id]=\Pim\,T\,A$. Indeed, the left hand side of this equation is
equal to $\Pim\,T\,(\lambda t.(A\,t)[\id])$ by definition, whereas the induction
hypothesis states that $(t:T)\ra (A\,t)[\id]=A\,t$.

\subsection{Relating the Syntax to a Model}
\label{sec:streicher}
It remains to show that the constructed syntactic model is initial. To achieve
this, we first define a relation between the syntactic model and an arbitrary
model, then show that the relation is functional, which lets us extract a
homomorphism from the relation.

This approach is an alternative version of Streicher's method for interpreting
preterms in an arbitrary model \cite{streichersemantics}. Streicher first defines
a family of partial maps from the presyntax to a model, then shows that the maps
are total on well-formed input. We have found that our approach is significantly
easier to formalise. To see why, note that the right notion of partial map in
type theory, which does not presume decidable definedness, is fairly
heavyweight:
\[
  \mathsf{PartialMap}\,A\,B := A\ra \big((P : \Set)\times\isaprop\,P\times (P \ra B)\big)
\]
In the above definition, we notice an opportunity for converting a fibered
definition of a type family into an indexed one; if we drop the propositionality
for $P$ for the time being, we may equivalently return a family indexed over
$B$, which is exactly just a relation $A\ra B\ra\Set$. Then, in our approach, we
recover uniqueness of $P$ through the functionality requirement on the $A\ra
B\ra\Set$ relation, and totality by already assuming well-formedness of $A$. In
type theory, using indexed families instead of display maps is a common
convenience, since the former are natively supported, while the
latter require carrying around auxiliary propositional equalities.

@@

Given a model $M$ of the QIIT, we define the functional relation satisfied by
the initial morphism $\rec : \mathsf{S}\to M$ by recursion on the typing judgments.
% As these judgments are propositions, we will sometimes
If $\GG$ is a context in $S$ and $\model{\GG}$ is a semantic context
(i.e.\ a context of the model $M$), we want to define a type
$\GG\sim \model{\GG}$ equivalent to $\rec\,\,\GG=\model{\GG}$. Of
course, at this stage, $\rec$ is not available yet since the point of
defining this relation is to construct $\rec$ in the end.

>>>>

We construct the theory of IIT signatures in the following steps:
\begin{enumerate}
\item We view the theory of signatures as a type theory, and we define
  its untyped syntax as mutual inductive types together with typing
  judgments given by inductive relations on the untyped syntax. Then
  the syntax $\I : \SignAlg$ is constructed using those untyped terms
  for which the typing relation holds.
\item We construct $\ll\blank\rr:\SignMor\,\I\,M$ for arbitrary
  $M : \SignAlg$, by:
  \begin{enumerate}
  \item defining a relation $\blank \sim \blank$ between the well-typed syntax
    and a given model. The idea is that given a syntactic context $\GG$ and a
    semantic context $\model{\GG}$ of the model $M$, we have
    $\GG \sim \model{\GG}$ if and only if $\ll\GG\rr
    = \model{\GG}$, and similarly for types, terms, and substitutions;
    % semantic context $\model{\GG}$ of
    % the model $M$ if and only if $\GG$ relates to $\rec\,{\GG}$;
    % enjoyed by the initial morphism from the
    % syntax to a given model;
  \item showing that this relation is functional.
    \end{enumerate}
  \item Proving the uniqueness of this morphism by showing that any morphism
    $f:\SignMor\,\I\,M$ satisfies the relation. For example, for any syntactic context
    $\GG$ we have $\GG\sim f\,\GG$.
      % TODO donner composantes sur les contextes
\end{enumerate}
The next sections detail each of these steps.

@@

\begin{definition}[Untyped syntax]
The untyped syntax is defined as the following inductive datatype.
\begin{alignat*}{5}
  & \rlap{(1) Substitution calculus} \\
  & \Conu && : \Set \\
  & \Tyu  && : \Set \\
  & \Subu  && : \Set \\
  % & \Subpre  && : \Set \\
  & \Tmu  && : \Set \\
  & \emptyu && : \Conu \\
    & \epsilonu && : \Subu \\
  & \blank\extu\blank && : \Conu\ra\Tyu\ra\Conu \\
  & \blank\consu\blank && : \Subu\ra\Tmu\ra\Subu \\
  & \varu  && : \N \ra \Tmu \\
  & \rlap{(2) Universe} \\
  & \Uu && : \Tyu \\
  & \Elu && : \Tmu\ra\Tyu \\
  & \rlap{(3) Inductive parameters} \\
  & \Piu && : \Tmu\ra\Tyu\ra\Tyu \\
  & \blank\appu\blank && : \Tmu\ra\Tmu\ra \Tmu \\
  & \rlap{(4) External parameters} \\
  & \Pimu && : (T:\Set)\ra(T\ra\Tyu)\ra\Tyu \\
  & \Piiu && : (T:\Set)\ra(T\ra\Tmu)\ra\Tmu \\
  & \blank\appmu \blank && : \Tmu\ra(\alpha:T) \ra\Tmu\\
  & \rlap{(5) Default value} \\
  & \erru && : \Tmu
\end{alignat*}
\end{definition}
Variables are modeled as de Bruijn indices, i.e.\ as natural numbers
pointing to a position in the context. We use the additional default
constructor $\erru:\Tmu$ in case of error (ill-scoped
substitution). The typing judgments will not mention $\erru$.  The
main interest of $\erru$ is that it behaves like a closed term (which
the theory of signatures lacks), in the sense that it is invariant
under substitution. This makes expected equalities about substitution
true even in the ill-typed case, thus reducing the number of
hypotheses for the corresponding lemmas (see
\Cref{lem:exch-laws}).

@@

\begin{definition}[Typing judgements]\label{def:typing_judgments}
  The typing judgments are defined as the following inductive type indexed over the
  untyped syntax:
\begin{alignat*}{5}
  & \rlap{(1) Substitution calculus} \\[0.5em]
  & \blank \vdash && : \Conu\ra \Set \\
  & \blank \vdash \blank  && : \Conu\ra \Tyu\ra\Set \\
  & \blank \vdash \blank \invar \blank  && : \Conu\ra \N \ra \Tyu\ra \Set \\
  & \blank \vdash \blank \in \blank  && : \Conu\ra \Tmu\ra \Tyu\ra \Set \\
  & \blank \vdash \blank \Rightarrow \blank  && : \Conu\ra \Subu \ra \Conu\ra \Set \\
  % & \Subpre  && : \Set \\
  & \emptyw && : \emptyu \vdash \\
  & \epsilonw && : \GG\vdash \epsilonu \Rightarrow \emptyu \\
  & \blank\extw\blank && : ( \GG\vdash )  \ra  (\GG\vdash A)  \ra
  \GG\extu A \vdash \\
  & \consw && :
    (\GD \vdash) \ra
    (\GG\vdash \sigma \Rightarrow\GD)\ra
    ( \GD \vdash A) \ra
    ( \GG \vdash t \in A[\sigma]) \ra
    \GG \vdash \sigma\consu t \Rightarrow \GD\extu A
   \\
  & \varw  && : (\GG \vdash n \invar A) \ra \GG \vdash \varu n \in A \\
  & \vzw  && : (\GG\vdash)\ra(\GG\vdash A)\ra\GG \extu  A \vdash \mathsf{0} \invar \wku \, A \\
  & \vsw  && : (\GG\vdash)\ra(\GG\vdash A)\ra(\GG \vdash n \invar A) \ra (\GG \vdash B) \ra \GG \extu  B
  \vdash \mathsf{S}\,n \invar \wku \, A \\
  & \rlap{(2) Universe} \\[0.5em]
  & \Uw && : (\GG \vdash)\ra \GG \vdash \Uu \\
  & \Elw && : (\GG \vdash)\ra (\GG \vdash a \in \Uu)\ra\GG \vdash \Elu\,a \\
  & \rlap{(3) Inductive parameters} \\[0.5em]
\  & \Piw && :
    (\GG \vdash)\ra (\GG \vdash a \in \Uu)\ra(\GG \extu
  \Elu\,a \vdash B)\ra\GG \vdash \Piu \, a \, B \\
  & \appw && :
    (\GG \vdash)\ra (\GG \vdash a \in \Uu)\ra(\GG \extu
    \Elu\,a \vdash B)
    \\
    & &&
    \ra(\GG \vdash t \in \Piu \, a \, B )
    \ra
    (\GG \vdash u \in \Elu\,a)
    \ra
    \GG \vdash t \appu u \in  B [0 := u] \\
  & \rlap{(4) External parameters} \\[0.5em]
  & \Pimw && :
    (T:\Set)\ra(A : T\ra \Tyu)\ra(\GG \vdash)\ra
    ((t : T)\ra\GG\vdash A\,t) \ra
    \GG \vdash \Pimu \,T\,A
    \\
    & \appmw  && :
    (T:\Set)\ra(A : T\ra \Tyu)\ra(\GG \vdash)\ra
    ((t : T)\ra\GG\vdash A\,t)
    \\
    & &&
    \ra(\GG \vdash t \in \Pimu \,T\,A)
    \ra (u : T) \ra \GG \vdash t\,\appmu\,u \in A\,u
  \end{alignat*}
\end{definition}
There is possibility of redundancy in the arguments of the
constructors. Here, we are ``{paranoid}'' (nomenclature from
\cite{paranoid}), so that we get more inductive hypotheses when
performing recursion.

@@

\begin{lemma}[Weakening preserves typing]\label{lem:weaken-typing}
% wkTelw : ∀ {Γp}{Ap}(Aw : Γp ⊢ Ap)Δp (Δw : (Γp ^^ Δp) ⊢) → ((Γp ▶p Ap) ^^ wkTel Δp) ⊢
% liftTw : ∀ {Γp}{Ap}(Aw : Γp ⊢ Ap)Δp{Bp}(Bw : (Γp ^^ Δp) ⊢ Bp) → ((Γp ▶p Ap) ^^ wkTel Δp) ⊢ (liftT ∣ Δp ∣ Bp)
\begin{alignat*}{5}
  & \wf{\wkO} && : (\GG \vdash A)\ra(\GG \merge \GD\vdash) \ra
  \GG \extu A \merge \wkO\, \GD \vdash \\
  % liftV (S (n + p)) (liftV n q) ≡ liftV n (liftV (n + p) q)
  & \wf{\wk} && : (\GG \vdash A)\ra(\GG \merge \GD\vdash B) \ra
  \GG \extu A \merge \wkO\,\GD \vdash \wk_{\length{\GD}}\, B  \\
  & \wf{\wk} && : (\GG \vdash A)\ra(\GG \merge \GD\vdash t\in B) \ra
  \GG \extu A \merge \wkO\,\GD \vdash \wk_{\length{\GD}}\, t \in \wk_{\length{\GD}}\,B  \\
  & \wf{\wkO} && : (\GG \vdash A)\ra(\GG \vdash \sigma
  \Rightarrow \GD) \ra
  \GG \extu A \vdash \wkO\, \sigma \Rightarrow \GD
  \end{alignat*}
\end{lemma}
\begin{proof}
By mutual induction on the typing relations.
\end{proof}

We show that judgments are stable under substitution.
\begin{lemma}[Substitution preserves typing]
\begin{alignat*}{5}
  % Tyw[] : ∀ {Γp}{Ap}(Aw : Γp ⊢ Ap) {Δp}(Δw : Δp ⊢){σp}(σw :  Δp ⊢ σp ⇒ Γp) → Δp ⊢ (Ap [ σp ]T)
  & \wf{[]} && : (\GG \vdash)\ra(\GD \vdash A)\ra
  (\GG \vdash \sigma \Rightarrow \GD) \ra
  \GG \vdash A [ \sigma ]
  \\
  & \wf{[]} && : (\GG \vdash)\ra(\GD \vdash t \in A)\ra
  (\GG \vdash \sigma \Rightarrow \GD) \ra
  \GG \vdash t [\sigma] \in A [ \sigma ]
  \\
  & \wf{[]} && : (\GD \vdash x \invar A)\ra
  (\GG \vdash \sigma \Rightarrow \GD) \ra
  \GG \vdash x [\sigma] \in A [ \sigma ]
  % ∘w : ∀ {Γ} {Δ σ} (σw :  Δ ⊢ σ ⇒ Γ)
  %  {Y}(Yw : Y ⊢) {δ} (δw :  Y ⊢ δ ⇒ Δ) →
  %  Y ⊢  (σ ∘p δ) ⇒ Γ
  \\
  & \wf{{\circ}} && :
  (\GG \vdash) \ra
  (\GG \vdash \sigma \Rightarrow \GD) \ra
  (\GD \vdash \tau \Rightarrow E) \ra
  \GG \vdash \tau \circ \sigma \Rightarrow E
  % keepw : ∀ {Γp}(Γw : Γp ⊢){Δp}(Δw : Δp ⊢){σp}(σw :  Γp ⊢ σp ⇒ Δp) {Ap}(Aw : Δp ⊢ Ap ∈ Up ) → (Γp ▶p (Elp Ap [ σp ]T )) ⊢ (keep σp) ⇒ (Δp ▶p Elp Ap)
  % & \wf{keep-\El} && : (\GG \vdash)\ra(\GD \vdash A)\ra
  % (\GG \vdash \sigma \Rightarrow \GD) \ra
  % \GG \vdash A [ \sigma ]
  % \\
\end{alignat*}
\end{lemma}
\begin{proof}
By mutual induction on the typing relations.
\end{proof}

We show the category and functor laws involving identity substitution
for well-formed types, terms and substitutions.
\begin{lemma}[Identity laws]\label{lem:identity}
\begin{alignat*}{5}
  % [idp]V : ∀ {Γ}{A}{x}(xw : Γ ⊢ x ∈v A) → (x [ idp ∣ Γ ∣ ]V) ≡ V x
  & [\idu] && : (\GG \vdash A)\ra A [ \idu\,\GG] = A \\
  & [\idu] && : (\GG \vdash x \invar A)\ra x [ \idu\,\GG] = V x \\
  & [\idu] && : (\GG \vdash t \in A)\ra t [ \idu\,\GG] = t \\
  & \idru && : (\GG \vdash \sigma \Rightarrow\GD)\ra \sigma \circ \idu\,\GG = \sigma \\
  & \idlu && : (\GG \vdash \sigma \Rightarrow\GD)\ra \idu\,\GD\circ \sigma = \sigma
  \end{alignat*}
\end{lemma}
Finally, we show that the identity substitution itself is well-typed:
\begin{lemma}[Typing for the identity substitution]
  \begin{alignat*}{5}
  & \idw && : (\GG \vdash)\ra \GG \vdash \idu\,\GG \Rightarrow \GG
  \end{alignat*}
\end{lemma}

\begin{definition}[Proposition]
  A type is a proposition, or proof-irrelevant, if it has at most one inhabitant.
  \[
    \isaprop\,T := (a : T) \ra (a' : T) \ra a = a'
  \]
\end{definition}

\begin{lemma}[Proof irrelevance of typing relations]\label{lem:uniq-typing-types}
\begin{alignat*}{5}
 & \isp{\Conw} && : \isaprop\, (\GG\vdash) \\
 & \isp{\Tyw} && : \isaprop\, (\GG\vdash A) \\
 & \isp{\wf\Var} && : \isaprop\, (\GG\vdash x \invar A) \\
 & \isp{\Tmw} && : \isaprop\, (\GG\vdash t \in A) \\
 & \isp{\wf\Sub} && : \isaprop\, (\GG\vdash \sigma \Rightarrow \GD)
\end{alignat*}
\end{lemma}

\begin{lemma}[Unicity of typing]
\begin{alignat*}{5}
  &
  \Tmw{=}\Ty & &:
  (\GG \vdash t \in A) \ra
  (\GG \vdash t \in B) \ra A = B
  \\
  &
  \wf\Var{=}\Ty & &:
  (\GG \vdash x \invar A) \ra
  (\GG \vdash x \invar B) \ra A = B
\end{alignat*}
\end{lemma}
Let us consider for instance the application constructor $\appw$: for a codomain
type $B$ it yields an overall type $C=B[\mathsf{0} := u]$ for an application. Even
if $C$ is known a priori, there may be another $B$ for which $B[\mathsf{0} := u] = C$,
possibly leading to many proofs that $t\appu u$ has type $C$. Unicity of typing
solves this issue, as $B$ is then uniquely determined by the type $\Piu\,A\,B$
of $t$.

@@

\subsubsection{The syntax as a signature algebra}

\begin{definition}[Syntax for the theory of signatures]\label{def:syntax}
  We define the syntax as an element of $\SignAlg$ by pairs of untyped
  syntax and typing relations:
% % \subsection{QIIT model}
% Then, we show that they are stable by weakening and substitution.
% Packing the untyped syntax with well-formed typed judgments yields
% a model $S$ of the QIIT. Let us give the most important components:
\begin{alignat*}{5}
 & \syn\Con && := \sum_\GG \GG\vdash
 \\
 & \syn\Ty \,(\GG,\wf\GG)&& := \sum_A \GG\vdash A
 \\
 & \syn\Tm \,(\GG,\wf\GG)(A,\wf A)&& := \sum_t \GG\vdash t \in A
 \\
 & \syn\Sub \,(\GG,\wf\GG)(\GD,\wf \GD)&& := \sum_\sigma \GG\vdash \sigma \Rightarrow \GD
\end{alignat*}
The other fields are given straightforwardly.  Regarding the equations, it is
enough to prove them only for the untyped syntactic part: as we argued in
\Cref{lem:uniq-typing-types}, the proofs of typing judgments are
automatically equal.
\end{definition}

\begin{remark}
  Up until \Cref{def:syntax}, UIP is not used. Function extensionality
  on the other hand is necessary because the untyped metatheoretic
  $\Pi$ takes a metatheoretic function as an argument. An example
  induction step that uses function extensionality is in
  \Cref{lem:identity}, in particular in the case
  $(\Pim\,T\, A)[\id]=\Pim\,T\,A$. Indeed, the left hand side of this
  equation is equal to $\Pim\,T\,(\lambda t.(A\,t)[\id])$ by
  definition, whereas the induction hypothesis states that
  $(t:T)\ra (A\,t)[\id]=A\,t$.
\end{remark}

\subsection{Relating the Syntax to a Model}
\label{sec:streicher}

It remains to show that the constructed syntax $\I$ is the initial
signature algebra. To achieve this, we first define a relation between
the syntax and any signature algebra, then show that the relation is
functional, which lets us extract a signature morphism from the relation.

This approach is an alternative presentation of Streicher's method for
interpreting preterms in an arbitrary model of type theory
\cite{streichersemantics}. Streicher first defines a family of partial
maps from the presyntax to a model, then shows that the maps are total
on well-formed input. We have found that our approach is significantly
easier to formalise. To see why, note that the right notion of partial
map in type theory, which does not presume decidable definedness, is
fairly heavyweight:
\[
  \mathsf{PartialMap}\,A\,B := A\ra \big((P : \Set)\times\isaprop\,P\times (P \ra B)\big)
\]
In the above definition, we notice an opportunity for converting a fibered
definition of a type family into an indexed one; if we drop the propositionality
for $P$ for the time being, we may equivalently return a family indexed over
$B$, which is exactly just a relation $A\ra B\ra\Set$. Then, in our approach, we
recover uniqueness of $P$ through the functionality requirement on the $A\ra
B\ra\Set$ relation, and totality by already assuming well-formedness of $A$. In
type theory, using indexed families instead of display maps is a common
convenience, since the former are natively supported, while the
latter require carrying around auxiliary propositional equalities.

@@

xxx

Given an $M : \SignAlg$, we define the functional relation satisfied by
the initial morphism $\rec : \mathsf{S}\to M$ by recursion on the typing judgments.
% As these judgments are propositions, we will sometimes
If $\GG$ is a context in $S$ and $\model{\GG}$ is a semantic context
(i.e.\ a context of the model $M$), we want to define a type
$\GG\sim \model{\GG}$ equivalent to $\rec\,\,\GG=\model{\GG}$. Of
course, at this stage, $\rec$ is not available yet since the point of
defining this relation is to construct $\rec$ in the end.
